---
title: "Fitting odin2 models to data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting odin2 models to data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

**WARNING**: The concepts here are stable, but some of the interface will change, in particular function names.

With a dynamical model we can simulate forwards in time and see how a system might change over time, given a set of parameters.  If we have a time series data set though, we can go a step further and find parameters consistent with the data.  This vignette gives an introduction to the approaches to fitting to data available for `odin2` models.  This support largely derives from the [`mcstate2`](https://mrc-ide.github.io/mcstate2) and [`dust2`](https://mrc-ide.github.io/dust2) packages and we will refer the reader to their documentation where further detail is on offer.

# Setting the scene

We'll start with a simple data set of daily cases of some disease over time

```{r}
data <- read.csv("incidence.csv")
head(data)
plot(cases ~ time, data, pch = 19, las = 1,
     xlab = "Time (days)", ylab = "Cases")
```

The data here shows a classic epidemic, with cases rising up to some peak and falling.  We will try fitting this with a simple compartmental SIR (Susceptible-Infected-Recovered) model, which we will write here in odin2.  There are a number of possible ways of writing this, but here we'll go for a stochastic discrete-time version, mostly because it will allow us to demonstrate a number of features of `odin2`, `dust2` and `mcstate2` (and because the ODE version is not yet written).

Before fitting the data, we'll write out a model that captures the core ideas (this is replicated from `vignette("odin2")`)

```{r}
sir <- odin2::odin({
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence) <- 0
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  update(incidence) <- if (time %% 1 == 0) n_SI else incidence + n_SI
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)
  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  beta <- parameter()
  gamma <- parameter()
  I0 <- parameter()
  N <- 1000
}, quiet = TRUE)
```

We can initialise this system and simulate it out over this time series and plot the results against the data:

```{r}
pars <- list(beta = 0.3, gamma = 0.1, I0 = 5)
sys <- dust2::dust_system_create(sir(), pars, n_particles = 20, dt = 0.25)
dust2::dust_system_set_state_initial(sys)
time <- 0:100
y <- dust2::dust_system_simulate(sys, time)
```

The `dust2::dust_system_simulate` function returns an `n_state` by `n_particle` by `n_time` matrix (here, 4 x 20 x 101).  We're interested in incidence (the fourth row here), and extracting that gives us a 20 x 101 matrix, which we'll transpose in order to plot it:

```{r}
matplot(time, t(y[4, , ]), type = "l", lty = 1, col = "#00000055",
        xlab = "Time (days)", ylab = "Cases", las = 1)
points(cases ~ time, data, pch = 19, col = "red")
```

The modelled trajectories are in grey, with the data points overlaid in red -- we're not doing a great job here of capturing the data.

# Comparing to data

We're interested in fitting this model to data, and the first thing we need is a measure of goodness of fit, which we can also code into the odin model, but first we'll explain the idea.

<!-- This should get more detailed coverage in dust docs, it would be good to link there once that is written -->

Our system moves forward in time until it finds a new data point; at this point in time we will have one or several particles present.  We then ask for _each_ particle how likely _this_ data point is.  This means that these calculations are per-particle and per-data-point.

Here, we'll use a [Poisson](https://en.wikipedia.org/wiki/Poisson_distribution) to ask "what is the probability of observing this many cases with a mean equal to our modelled number of daily cases".

The syntax for this looks a bit different to the odin code above:

```{r}
sir <- odin2::odin({
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence) <- 0
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  update(incidence) <- if (time %% 1 == 0) n_SI else incidence + n_SI
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)
  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  beta <- parameter()
  gamma <- parameter()
  I0 <- parameter()
  N <- 1000

  # Comparison to data
  cases <- data()
  compare(cases) ~ Poisson(incidence)
}, quiet = TRUE)
```

These last two lines are the new addition to the odin code.  The first says that `cases` will be found in the data.  The second restates our aim from the previous paragraph, comparing the observed `cases` against modelled `incidence`.  The syntax here is designed to echo that of the [`mcstate2` DSL](https://mrc-ide.github.io/mcstate2/articles/dsl.html), though (at least for now) we use `compare()` on the left side to make this clearer.

With this version of the model we can compute likelihoods with `dust2`'s machinery.

# Stochastic likelihood with a particle filter

Our system is **stochastic**; each particle will produce a different trajectory and from that a different likelihood.  Each time we run the system we get a different combination of likelihoods.  We can use a **particle filter** to generate an estimate of the marginal likelihood, averaging over this stochasticity.  This works by resampling the particles at each point along the time series, according to how likely they are.

```{r}
filter <- dust2::dust_filter_create(sir(), 0, data, n_particles = 200)
```

Each time we run this filter the likelihood will be slightly (or very) different:

```{r}
dust2::dust_filter_run(filter, pars)
dust2::dust_filter_run(filter, pars)
```

If you run the filter enough times a distribution will emerge of course.  Let's compare two points in parameter space, varying the `beta` parameter and running the filter 100 times each:

```{r}
pars1 <- modifyList(pars, list(beta = 0.25))
pars2 <- modifyList(pars, list(beta = 0.23))
ll1 <- replicate(100, dust2::dust_filter_run(filter, pars1))
ll2 <- replicate(100, dust2::dust_filter_run(filter, pars2))

xb <- seq(floor(min(ll1, ll2)), ceiling(max(ll1, ll2)), by = 1)
hist(ll2, breaks = xb, col = "#0000ff99", freq = FALSE,
     xlab = "Log likelihood", ylab = "Density", main = "")
hist(ll1, breaks = xb, add = TRUE, freq = FALSE, col = "#ff000099")
abline(v = c(mean(ll1), mean(ll2)), col = c("red", "blue"), lwd = 2)
```

So even a relatively small difference in a parameter leads to a difference in the log-likelihood that is easily detectable in only 100 runs of the filter, even when the distributions overlap.  However, it does make optimisation-based approaches to inference, such as maximum likelihood, tricky because it's hard to know which way "up" is if each time you try a point it might return a different height.

If you run a particle filter with the argument `save_history = TRUE` then we save the trajectories of particles over time:

```{r}
dust2::dust_filter_run(filter, list(beta = 0.2, gamma = 0.1), save_history = TRUE)
```

You can access these with `dust2::dust_filter_last_history()`:

```{r}
h <- dust2::dust_filter_last_history(filter)
```

The result here is a 4 x x 100 200 array:

```{r}
dim(h)
```

The dimensions represent, in turn:

1. 4 state variables
2. 200 particles
3. 100 time steps (corresponding to the data)

Considering just the fourth state, representing incidence, and plotting over time, you may be able to make out the tree structure of the trajectories, with fewer distinct traces at the start of the time series, and some traces more heavily represented in the final sample than others:

```{r}
matplot(t(h[4, , ]), type = "l", lty = 1, col = "#00000022")
points(cases ~ time, data, pch = 19, col = "red")
```

# Inference with particle MCMC (pMCMC)

We can use MCMC to explore this model, but to do this we will need a prior.  We'll use [mcstate2's DSL](https://mrc-ide.github.io/mcstate2/articles/dsl.html) to create one; this looks similar to the odin code above:

```{r}
prior <- mcstate2::mcstate_dsl({
  beta ~ Exponential(mean = 0.3)
  gamma ~ Exponential(mean = 0.1)
})
```

Here we define a prior that covers `beta` and `gamma`, two of the three input parameters to our odin model.  This prior is an `mcstate_model` object, which we can use to sample from, compute log densities with (to compute the prior), etc.

We also need to adapt our `dust2` filter object above for use with `mcstate2`.  All we need to do here is to describe how a vector of statistical parameters (here `beta` and `gamma`) will be converted into the inputs that the `sir` system needs to run (here a list with elements `beta`, `gamma` and `I0`).  We do this with an `mcstate2::mcstate_packer` object:

```{r}
sir_packer <- mcstate2::mcstate_packer(c("beta", "gamma"), fixed = list(I0 = 5))
```

With this packer we can convert from a **list** of name-value pairs suitable for initialising an `odin2` model into a **vector** of parameters suitable for use with `mcstate2`:

```{r}
sir_packer$pack(pars)
```

and we can carry out the inverse:

```{r}
sir_packer$unpack(c(0.3, 0.1))
```

Combining the filter and packer we create an `mcstate2` model, which we'll call `likelihood`, as that's what it represents:

```{r}
likelihood <- dust2::dust_filter_mcstate(filter, sir_packer)
```

This likelihood is now also an `mcstate_model` object:

```{r}
likelihood
```

The `mcstate2` package provides a high-level interface for working with these objects.  For example, to compute the likelihood we could now use `mcstate2::mcstate_model_density()`:

```{r}
mcstate2::mcstate_model_density(likelihood, c(0.2, 0.1))
```

The difference to using `dust2::dust_filter_run` here is now we provide a parameter vector from our *statistical model*, rather than the inputs to the odin/dust model.  This conforms to the interface that `mcstate2` uses and lets us run things like MCMC.

We can combine the prior and the likelihood to create a posterior:

```{r}
posterior <- prior + likelihood
```

The last ingredient required for running an MCMC is a sampler.  We don't have much choice with a model where the likelihood is stochastic, we'll need to run a simple random walk. However, for this we still need a proposal matrix (the variance covariance matrix that is the parameter for a multivariate Gaussian - we'll draw new positions from this).  In an ideal world, this distribution will have a similar shape to the target distribution (the posterior) as this will help with mixing.  To get started, we'll use an uncorrelated random walk with each parameter having a fairly wide variance of 0.02

```{r}
sampler <- mcstate2::mcstate_sampler_random_walk(diag(2) * 0.02)
```

We can now run an MCMC for 100 samples

```{r}
samples <- mcstate2::mcstate_sample(posterior, sampler, 100, 
                                    initial = sir_packer$pack(pars))
```

We need to develop nice tools for working with the outputs of the sampler, but for now bear with some grubby base R manipulation.

The likelihood here is very "sticky"

```{r}
plot(samples$density, type = "l")
```

It's hard to say a great deal about the parameters `beta` (per-contact transmission rate) and `gamma` (recovery rate) from this few samples, especially as we have very few *effective* samples:

```{r}
plot(t(drop(samples$pars)), pch = 19, col = "#00000055")
```

## Effective sampling

There are several things we can do here to improve how this chain mixes

* We can try and find a better proposal kernel.
* We can increase the number of particles used in the filter.  This will reduce the variance in the estimate of the marginal likelihood, which means that the random walk will be less confused by fluctuations in the surface it's moving over.  This comes at a computational cost though.
* We can increase the number of threads (effectively CPU cores) that we are using while computing the likelihood.  This will scale fairly efficiently through to at least 10 cores, with the likelihood calculations being almost [embarrassingly parallel](https://en.wikipedia.org/wiki/Embarrassingly_parallel).  This will help to offset some of the costs incurred above.
* We can run multiple chains at once.  We don't yet have a good parallel runner implemented in `mcstate2` but it is coming soon.  This will reduce wall time (because each chain runs at the same time) and also allows us to compute convergence diagnostics which will reveal how well (or badly) we are doing.
* We can try a deterministic model (see below) to get a sense of the general region of high probability space.

Here, we apply most of these suggestions at once, using a variance-covariance matrix [that I prepared earlier](https://www.youtube.com/watch?v=ziqD1xvSpF4):

```{r}
filter <- dust2::dust_filter_create(sir(), 0, data, n_particles = 1000)
likelihood <- dust2::dust_filter_mcstate(filter, sir_packer)
vcv <- matrix(c(0.0005, 0.0003, 0.0003, 0.0003), 2, 2)
sampler <- mcstate2::mcstate_sampler_random_walk(vcv)
samples <- mcstate2::mcstate_sample(posterior, sampler, 1000, 
                                    initial = sir_packer$pack(pars))
```

The likelihood now quickly rises up to a stable range and is clearly mixing:

```{r}
plot(samples$density, type = "l")
```

The parameters `beta` (per-contact transmission rate) and `gamma` (recovery rate) are strongly correlated

```{r}
plot(t(drop(samples$pars)), pch = 19, col = "#00000055")
```

# Deterministic models from stochastic

Another way of fitting this model is to simply throw away the stochasticity.  In the model above we have the lines

```r
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)
```

which are the stochastic portion of the model.  Each time step we compute the number of individuals who make the transition from `S` to `I` and from `I` to `R` by sampling from the binomial distribution.  We can replace these calls by their expectations (so effectively making `n_SI = S * p_SI`) by running the model in "deterministic" mode.

This simplification of the stochastic model can be seen as taking expectations of the underlying random process, but there's no reason to expect that this represents the mean of the whole model ($E[f(x)] \neq f(E[x])$, at least generally).

We have found these simplifications useful:

* They are not stochastic, so you can use adaptive MCMC or other more efficient algorithms
* They are orders of magnitude faster, because instead of running 100s or thousands of particles per likelihood evaluation you just run one
* The region of high probability density of the deterministic model is often within the (broader) region of high probability density of the stochastic model, so you can use these models to create reasonable starting parameter values for your chains
* The signs and relative magnitudes of the covariances among parameters are often similar between the deterministic and stochastic model, so you can use the deterministic model to estimate a variance-covariance matrix for your stochastic model -- though you will need to increase all quantities in it

Obviously, this approximation comes with costs though:

* You no longer have integer valued quantities from the expectations of samples in your discrete distributions, so you have to think about fractional individuals
* The model can no longer account for stochastic effects, e.g., at low population sizes.  This can make the model overly rigid, and it may poorly account for observed patterns
* The fixed `dt` approach is a [first order Euler solver](https://en.wikipedia.org/wiki/Euler_method) which offers few stability guarantees, and this will differ from a system of ODEs solved with a better ODE solver

To create a deterministic "filter" (currently, and temporarily called an "unfilter"), use `dust_unfilter_create()` in place of `dust_filter_create`.  This will replace all calls to stochastic functions with their expectations at the point of call.

```{r}
unfilter <- dust2::dust_unfilter_create(sir(), 0, data)
```

In contrast with `filter`, above, multiple calls to `unfilter` with the same parameter set yield the same result.

```{r}
dust2::dust_unfilter_run(unfilter, pars)
dust2::dust_unfilter_run(unfilter, pars)
```

We can now proceed as before, reusing our `packer`, `prior` and `sampler` objects, which are still useable here:

```{r}
likelihood_det <- dust2::dust_filter_mcstate(unfilter, sir_packer)
posterior_det <- prior + likelihood_det
samples_det <- mcstate2::mcstate_sample(posterior_det, sampler, 1000,
                                        initial = sir_packer$pack(pars))
```

Here, you can see the 1000 samples from the deterministic model (in blue) overlaid on top of the samples from the stochastic model (in grey):

```{r}
plot(t(drop(samples$pars)), pch = 19, col = "#00000033")
points(t(drop(samples_det$pars)), pch = 19, col = "#0000ff33")
```

The estimated parameters here look overall shifted higher in the deterministic model, and the correlation between the parameters stronger.  However, if we had no idea about what "good" parameters might be, this can get us into the approximately right location.

# Differentiable models

We can go a step further than simply turning off stochasticity to create a deterministic model; now that we've got a deterministic likelihood function we can also differentiate that likelihood with respect to (some of) the parameters.

```{r}
sir <- odin2::odin({
  initial(S) <- N - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(incidence) <- 0
  update(S) <- S - n_SI
  update(I) <- I + n_SI - n_IR
  update(R) <- R + n_IR
  update(incidence) <- if (time %% 1 == 0) n_SI else incidence + n_SI
  n_SI <- Binomial(S, p_SI)
  n_IR <- Binomial(I, p_IR)
  p_SI <- 1 - exp(-beta * I / N * dt)
  p_IR <- 1 - exp(-gamma * dt)
  beta <- parameter(differentiate = TRUE)
  gamma <- parameter(differentiate = TRUE)
  I0 <- parameter()
  N <- 1000

  # Comparison to data
  cases <- data()
  compare(cases) ~ Poisson(incidence)
}, quiet = TRUE)
```

This the same model as above, except for the definition of `beta` and `gamma`, which now contain the argument `derivative = TRUE`.

This system can be used as a stochastic model (created via `dust2::dust_filter_create`) just as before.  The only difference is where the model is created using `dust2::dust_unfilter_create()`.

```{r}
unfilter <- dust2::dust_unfilter_create(sir(), 0, data)
```

When you run the unfilter, you can now provide the argument `adjoint = TRUE` which will enable use of `dust_unfilter_last_gradient()` (we may make this the default in future).

```{r}
dust2::dust_unfilter_run(unfilter, pars, adjoint = TRUE)
dust2::dust_unfilter_last_gradient(unfilter)
```

We can create an `mcstate2` model with this, as before:

```{r}
likelihood <- dust2::dust_filter_mcstate(unfilter, sir_packer)
likelihood
```

and this model advertises that it can compute gradients now!

So from `mcstate2` we can use `mcstate2::mcstate_model_density()` and `mcstate2::mcstate_model_gradient()` to compute log-likelihoods and gradients.

```{r}
mcstate2::mcstate_model_density(likelihood, c(0.2, 0.1))
mcstate2::mcstate_model_gradient(likelihood, c(0.2, 0.1))
```

Because the prior contained gradient information, a posterior created with this version of the model also has gradients:

```{r}
posterior <- likelihood + prior
posterior
```

With a model configured this way, you can use the [Hamiltonian Monte Carlo](https://en.wikipedia.org/wiki/Hamiltonian_Monte_Carlo) method with `mcstate2::mcstate_sampler_hmc()`, which can be far more efficient than a random walk once tuned.

**WARNING**": Using `derivative = TRUE` on some parameters has the effect of making the rest use `constant = TRUE`.  We will describe the effects of this in a vignette on differentiable models, soon.

# Running multiple parameter sets at once

You can efficiently run multiple parameter sets at once; this will be parallelised where possible when enabled.

## For dust2 systems

When initialising the dust system, you should:

* pass an (typically unnamed) list of parameters, each element of which is a different set of parameters for the system
* pass the `n_groups` argument indicating how many groups you wish to initialise

Here is a simple case with two parameter sets that differ in `beta`, each run with 20 particles:

```{r}
pars2 <- list(list(beta = 0.3, gamma = 0.1, I0 = 5),
              list(beta = 0.2, gamma = 0.1, I0 = 5))
sys <- dust2::dust_system_create(sir(), pars2, n_particles = 20, n_groups = 2,
                                 dt = 0.25)
dust2::dust_system_set_state_initial(sys)
time <- 0:100
y <- dust2::dust_system_simulate(sys, time)
```

The dimensions of `y` is now

```{r}
dim(y)
```

representing

* 4 state variables
* 20 particles
* 2 parameter groups
* 101 times

Consider just incidence as above:

```{r}
matplot(time, t(y[4, , 1, ]), type = "l", lty = 1, col = "#ff000055",
        xlab = "Time (days)", ylab = "Cases", las = 1)
matlines(time, t(y[4, , 2, ]), type = "l", lty = 1, col = "#0000ff55")
points(cases ~ time, data, pch = 19)
```

## For dust2 filters/unfilters

Here we assume (require, really) that each parameter set is associated with a different data set.  We may relax this in future, but this is the typical use case we have seen.  We need an additional column called `group` in addition to `time`:

```{r, include = FALSE}
data2 <- local({
  d1 <- data
  d2 <- data
  d2$cases <- rpois(nrow(d2), d2$cases * 1.5)
  data2 <- cbind(group = rep(1:2, each = nrow(d2)), rbind(d1, d2))
  data2 <- data2[order(data2$time), ]
  rownames(data2) <- NULL
  data2
})
```

```{r}
head(data2)
```

(this is just synthetic data for now, created by duplicating and perturbing the original data).

```
plot(cases ~ time, data2, subset = group == 2, pch = 19, col = "red",
     xlab = "Time (days)", ylab = "Cases")
points(cases ~ time, data2, subset = group == 1, pch = 19, col = "blue")
```

Because the data is grouped, we don't need to tell `dust2::dust_filter_create()` that we have two groups, though you can pass `n_groups = 2` here if you want, which will validate that you really do have exactly two groups in the data:

```{r}
filter2 <- dust2::dust_filter_create(sir(), 0, data2, n_particles = 200)
```

When passing parameters into the filter, you now should mirror the format used in `dust2::dust_system_run()`; a list of lists:

```{r}
dust2::dust_filter_run(filter2, pars2)
```

We now have two likelihoods returned by the filter; one per group.

For the deterministic unfilter the process is the same:

```{r}
unfilter2 <- dust2::dust_unfilter_create(sir(), 0, data2)
dust2::dust_unfilter_run(unfilter2, pars2)
```

however, our gradient has picked up a dimension:

```{r}
dust2::dust_unfilter_last_gradient(unfilter2)
```

Here, the first **column** is the gradient of the first parameter set, and the first **row** is the gradient of `beta` over parameter sets.

Compare with the single parameter case:

```{r}
dust2::dust_unfilter_run(unfilter, pars2[[1]])
dust2::dust_unfilter_last_gradient(unfilter)
```

## For mcstate2 models

Here, rather than a parameter vector you would pass in a matrix, where each column represents a parameter set:

```{r}
p <- cbind(c(0.2, 0.1), c(0.25, 0.1))
```

However, there's a small bit of functionality we need to add here in dust2, still.

<!-- ```{r,} -->
<!-- likelihood2 <- dust2::dust_filter_mcstate(filter2, sir_packer) -->
<!-- mcstate2::mcstate_model_density(likelihood2, p) -->
<!-- ``` -->

# Further reading

This vignette can only skim the surface, and is organised around features of odin itself.  The machinery for running the models comes from [`dust2`](https://mrc-ide.github.io/dust2) and for performing inference from [`mcstate2`](https://mrc-ide.github.io/mcstate2) and we will be adding documentation to those packages that covers the details of use.
