---
title: "Migrating from odin 1.x.x"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from odin 1.x.x}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("helpers.R")
fast_and_quiet_dust()
```

This vignette discusses key differences with odin version 1, and with `odin.dust`.

```{r}
library(odin2)
library(dust2)
```

# New features

Some of these features were present in versions of `odin.dust` and many derive from underlying support in `dust2`.

* Comparison to data and likelihood support (introduced in `odin.dust`)
* Automatic differentiation
* More efficient setting of the subset of parameters you are likely to use while fitting (use the `constant` argument to `parameter()`)
* Multiple parameter sets at once (introduced in `odin.dust` but expanded here)
* Run multiple copies of a system at once in parallel (introduced in `odin.dust`)
* Built-in support for periodic variable resetting (e.g., for computing daily incidence)
* Better (we hope) error messages
* Better debugging tools (see `vignette("debugging")`)
* Compile-time bounds checking of arrays, preventing many crashes

## Planned

* Optional array bounds checking, during runtime (in the latter case with a performance penalty once enabled).

## Hoped

* Support for multinomial samples and other vector-valued functions

# Missing features

Things we do plan on implementing:

* Compile-time parameter substitution (`mrc-5575`)
* Compilation to JavaScript
* Compilation to GPU

Things that we plan to drop in this version

* Many details in `config()` and `options`

Note that some errors are still not caught as odin errors, and invalid odin code will be accepted and generate C++ code that fails to compile.  Please send us code that causes this to happen.

# Changes in syntax

## `user()` becomes `parameter()`

This might be the largest user-visible change, and can be automatically migrated.

Previously, to support parameters you might write

```r
a <- user(4)
```

which says that `a` is a user-supplied parameter with a default value of `4`.  In most cases this now simply becomes

```r
a <- parameter(4)
```

The `integer` argument accepted by `user` has now changed:

* `user(integer = TRUE)` becomes `parameter(type = "integer")`
* `user(integer = FALSE)` becomes `parameter(type = "real")`

This translation can be done automatically in most cases, and will be done (with a warning) by default if possible.  You should update your code with the suggested fix, however, as this translation will be removed in a future version.

## Compare keyword is now removed.

In comparisons such as

```r
compare(d) ~ Normal(0, 1)
```

The `compare` keyword, and the `~` only occur together.  This has been simplified, and is now written as:

```r
d ~ Normal(0, 1)
```

which reads as: `d` is normally distributed with a mean of 0 and standard deviation of 1.

## Vector parameters assign without array indices

Previously, if you had a vector parameter you had to write

```r
a[] <- parameter()
dim(a) <- 10
```

(though with `user()`, as in the previous section).  However, the array index here does not really add anything as we already know how many dimensions `a` has from the `dim` call.  So now you should write

```r
a <- parameter()
dim(a) <- 10
```

which makes it clearer that **all** of `a` is assigned by the parameter call.

## Vector/matrix/array parameters whose size is determined by input require `rank` argument

What a mouthful.  Previously you might have written

```r
a[, ] <- user()
dim(a) <- user()
```

which means "`a` is a matrix whose dimensions are determined by the input we are given on initialisation".  Because of the previous change the first line changes to

```r
a <- parameter()
```

but that means that we no longer know that `a` has two dimensions.  That's ok because we've moved the responsibility for this into the `dim()` assignment line anyway (internally).  So for now you write

```r
dim(a) <- parameter(rank = 2)
```

which conveys the same intent.  We may make this slightly more friendly in future (see `vignette("functions")`).

## Interpolate results assign without array indices

Previously, if you had an `interpolate()` call that returned a vector (or higher-dimension array) you had to write

```r
v[] <- interpolate(a, b, "constant")
```

but now you should drop the `[]`, as for the `parameter()` case above, as you are replacing *all* of `v` at once, writing:

```r
v <- interpolate(a, b, "constant")
```

## Discrete-time models have a more solid time basis

Previously, discrete time models used `step` to count steps forward as unsigned integers, usually from zero.  Many models added a parameter (or constant) `dt` representing the timestep and then a variable `time` which represented the time as a real-valued number.  For example you might have `dt` of 0.25 and then your model stops at times `[0, 0.25, 0.5, 0.75, 1]` for steps `[0, 1, 2, 3, 4]`.

We formalise this approach now having discrete time systems be explicitly in terms of the same time basis as ODE models (that is, some real valued time axis).  When you initialise a model you pass in `dt`, which must be an even divisor of 1 (so 0.5, 0.25, 0.2, etc).  We then take steps of this size.  The wrinkle is that (at least for now) the model will only return control back to you, or state back to you, at integer-valued times.  We may relax this in future to allow returning at any time value that is a multiple of `dt`.

This will cause a few issues for using old code, which we cover below.

### Assignments to `dt`

You may have models that assign to `dt`, either directly or as a parameter.  You can no longer do this as `dt` will be provided by `dust` (see `dust2::dust_system_create()`).

We can automatically remove these (with a warning) in some cases.

### Assignments to `time`

Conventionally, many models would write

```r
time <- step * dt
```

which is the linear transformation of time that dust2 now does.  We can remove these statements and your model should work as intended.

### Use of `step`

All other uses of `step` are problematic and will need manual fixing.  We will try and accumulate migration strategies here, so please let us know if you have had to do anything not listed.

**Access "interpolated" values from a grid**: In `sircovid` we used `step` as an array index, in order to support time-varying inputs (e.g., vaccine allocation schedules, rates of contact).  This is no longer supported (at all) because `dt` is changed separately from the inputs.  Instead you should use odin's interpolation functions.

**Periodic resetting**: You may have written:

```
a <- if (step %% freq == 0) b else c
```

to have some quantity that took different values every `freq` steps, where `freq` is usually `1/dt` or `m/dt` where `m` is an integer.  You should rewrite this to use `time`:

```
a <- if (time * dt / m == 0) b else c
```

## The name of the time variable in continuous time models has changed

Previously, time was `t` but we have moved this to `time` to be a little more explicit.  We can automatically migrate your code in many cases, unless you have defined a variable `time` already.

## Random number function calls have changed

Previously we used the same names as R's random-number-drawing functions, for example `rbinom` for drawing from a binomial distribution.  This has changed to use the distribution name instead.

The motivating reason for this change was that in odin we might write

```r
rbinom(size, prob)
```

but if you were writing this in R you would write

```
rbinom(1, size, prob)
```

with the first argument being the number of draws from the distribution in question.  This departure in arguments feels needlessly confusing!  If you were using `odin` without `odin.dust` then this did compile to a call to one of R's underlying random number functions so this connection was reasonable but from version 2 we use monty's parallelisable random number distributions.

The mapping is:

* `rbeta()` to `Beta`
* `rbinom()` to `Binomial`
* `rcauchy()` to `Cauchy`
* `rchisq()` to `ChiSquared` (unsupported for now)
* `rexp()` to `Exponential`
* `rf()` to `F` (unsupported for now)
* `rgamma()` to `Gamma`
* `rgeometric()` to `Geometric` (unsupported for now)
* `rhyper()` to `Hypergeometric`
* `rlogis()` to `Logistic` (unsupported for now)
* `rlnorm()` to `LogNormal`
* `rnbinom()` to `NegativeBinomial`
* `rnorm()` to `Normal`
* `rpois()` to `Poisson`
* `rt()` to `T` (unsupported for now)
* `runif()` to `Uniform`
* `rweibull()` to `Weibull`

(Not all of these are implemented yet).

## System size cannot be changed after creation

This limitation comes from our implementation in `dust2` and it is possible to relax it in some settings.  However, it is fairly important for efficiently running the system within a pMCMC context where we save state periodically.

If your system has a parameter that affects the number of state variables in the system (e.g., the number of age categories that a compartment is stratified by), you may not change this after initialisation.  This will be prevented by the parser once arrays are implemented.

## Changes in the way arrays are handled

The two-argument form of `dim()` has been removed, as we did not believe it was used and it is confusingly different to R.  Previously you could write `dim(x, 3)` to get the length of the third dimension of `x`; this is no longer supported.  Please let us know if this is a problem.

# Changes to how delays are supported

We support, to a degree, delay differential equations.  That they work at all is a minor miracle, especially given that the solver we use (in dust2) is extremely simple.

## `delay()` cannot be used in discrete time models

This was removed in `odin.dust` and the limitation remains, at least for now.  We may enable this later if required, please let us know.  The reason for this change is that the way that delays are handled in continuous time is fundamentally different to how we would do this in discrete time (particularly with stochastic expressions), but we do not have a good motivating reason to develop this.

## `delay()` results assign without array indices

Previously, if you had a `delay()` call that returned a vector (or higher-dimension array) you had to write

```r
v[] <- delay(a, tau)
```

but now you should drop the `[]`, as for the `parameter()` and `interpolate()` cases above, as you are replacing *all* of `v` at once, writing:

```r
v <- delay(a, tau)
```

## The first argument to `delay()` must be the name of an equation

Previously, you could delay an expression (e.g., `x <- delay(a + b, 1)`).  For simplicity in implementation we have dropped this and require that the first argument is a symbol.  We might change this later, but in the current version it makes the syntax around handling delayed arrays a little more consistent (see above).

## The 'default' argument to `delay()` has been removed

Previously, you could write

```r
x <- delay(y, tau, default)
```

to provide a default argument to the delayed value before `tau` time had elapsed.  This was not widely advertised or used, and has been removed.

## The delay time must be constant

We require that the delay time is known at first model initialisation and not updated subsequently (i.e., that it is numeric, a literal value assigned to symbol or a paramter with `constant = TRUE`).

# General changes

This package replaces `odin.dust` and will eventually replace `odin` (as in, we'll copy the entire `odin2` code into `odin` to become version 2.0.0 of that package).

The relationship between packages has changed.  Previously `mcstate` "knew" about `dust` models and so you had to use `odin.dust` practically to use the statistical machinery in `mcstate`.  We've changed this around now, so that `odin2` "knows" about `monty` and can create systems that will work well with `monty`.  We now depend on `monty`, so if you have `odin2` installed you can start working towards fitting models immediately.

# Known limitations

## Much slower compilation time

Because we now compile to C++ via `dust2`, the compilation times have massively increased.  Previously, compilation of a simple model took less than a second, but now this will take 6 seconds or so.  You can alleviate this to a degree during development by specifying `debug = TRUE` when compiling, which reduces this down to about 3 seconds.  These times are from my workstation but I expect the relative differences to hold (we're probably 10x slower than previously but can be "only" 5x slower if you turn off optimisation).  If you were previously using `odin.dust` you should notice little change here.

## Loss of features from odin 1.x

Some original odin 1.x features have been lost:

For discrete-time systems:

* **You may not use `output()`**; see below for workarounds and the reasons for this change.
* **Delays are not supported**; they never worked well, particularly for stochastic systems.

# Updating old code

If you compile odin code that contains any of the changes above, it will try and update the code to the new version and keep going:

```{r}
gen <- odin2::odin({
  initial(x) <- 1
  deriv(x) <- x + a / t
  a <- user(2)
})
```

This model contains two issues that can be easily rewritten; the solution to this rewriting is printed to screen and the model is compiled as if you had rewritten them.

Not everything can be rewritten, especially changes involving `step`:

```{r, error = TRUE, purl = FALSE}
gen <- odin2::odin({
  initial(x) <- 1
  update(x) <- x + a / step
  a <- user(2)
})
```

In this case, odin errors and tries to indicate where you have work to do (and directs you to this document!)

For code saved into a file, you can use `odin_migrate` to migrate code from the old syntax to the new; this will preserve comments and formatting except for code that is rewritten so it should be fairly unintrusive.

```{r, include = FALSE}
path <- tempfile(fileext = ".R")
writeLines(c(
  "initial(x) <- 1",
  "deriv(x) <- x + a / t",
  "a <- user(2)"),
  path)
```

For example, in `path` (a temporary file for this vignette) we have saved the code from above:

```{r echo = FALSE, results = "asis"}
r_output(readLines(path))
```

We can migrate this in-place with:

```{r}
odin_migrate(path, path)
```

and now the code contains:

```{r echo = FALSE, results = "asis"}
r_output(readLines(path))
```

## Avoiding `output()` in discrete-time systems

If you have odin 1.x code, you may have a system that uses `output()` in discrete-time; this will fail to compile.

```{r, error = TRUE}
odin({
  initial(x) <- 1
  update(x) <- x + 1
  output(y) <- x / 2
})
```

Previously, this would have worked to create a system where you have one input variable (on which the system depends) and one output variable (entirely derived from the inputs), producing output like

```
#>      step x   y
#> [1,]    0 1 0.5
#> [2,]    1 2 1.0
#> [3,]    2 3 1.5
#> [4,]    3 4 2.0
#> [5,]    4 5 2.5
#> [6,]    5 6 3.0
```

(this comes from [the `odin.dust` migration guide](https://mrc-ide.github.io/odin.dust/articles/porting.html#avoiding-output))

This is important in ODE models because there are often things that you want to observe that are functions of the system but for which you can't write out equations to describe them in terms of their rates --- the sum over a set of variables for example.

Note here how the `x` used in the calculation is really the `x` at the *end* of a time step, not the `x` on inputs, which means that the output column satisfies `y == x / 2`.  This turns out to be very hard to get right and reason about, and involved some fairly unpleasant bookkeeping in `dde` (the package that we used to drive these systems); in effect we run an additional time step at the end of the run in order to compute these `output` variables, and that is inefficient for `dust`'s use within a particle filter, and poorly behaved with anything that used random numbers.

Additionally, there is no great need for `output` for discrete time models, as we can treat `y` above as just another variable.  This also allows us to be more explicit about when within the time step this output is being computed:

```{r}
gen <- odin({
  initial(x) <- 1
  new_x <- x + 1
  update(x) <- new_x
  initial(y1) <- 0
  initial(y2) <- 0
  update(y1) <- x / 2
  update(y2) <- new_x / 2
})
```

We can run this:

```{r}
sys <- dust_system_create(gen)
y <- dust_system_set_state_initial(sys)
y <- dust_system_simulate(sys, 0:5)
dust_unpack_state(sys, y)
```

Here, the variable `x` is updated as for the original output. The variable `y1` column computes the relationship with the `x` at the beginning of the time step, while `y2` is most equivalent to our previous `output` command and requires storing the that will become the updated `x` in order to use that value to both update `x` and `y2`.
